---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
```

```{r packages}
require(tidyverse)
```

# Title Page

Include your names

Include the tutorial group number

Include your tutorial lecturer's name

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
dataset <- midwest
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.


# ---- PACKAGES ----
install.packages("tidyverse")
install.packages("dplyr")
install.packages("ggplot2")
install.packages("scales")
install.packages("sf") 
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")
install.packages("viridis")
library(scales)
library(tidyr)
library(dplyr)
library(readxl)
library(ggplot2)
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
theme_set(theme_bw())

# ---- DATA CLEANING EN MERGING ----


#Dataframe maken van excel-bestand
df1 <- read_excel("Studieschuld_2011-2024(goeie).xlsx", skip = 1)
df2 <- read_excel("England_figure_10.xlsx")
df3 <- read_excel("Average Total Student Debt USA.xlsx")



#Rijen en Kolommen die niet nodig zijn verwijderen
df1 <- df1[-c(1, 2, 3), ]
df1 <- df1[, -c(2, 3, 6)]
df1 <- df1[, -c(2) ]
df2 <- df2[, -c(3)]


#Veranderen van schooljaar naar kalenderjaar
df2$'Financial year' <- 2007:2024  


#Kolomnamen overal hetzelfde maken voor samenvoegen
colnames(df1)[1] <- "Year"
colnames(df1)[2] <- "Studieschuld_NL"
colnames(df2)[1] <- "Year"
colnames(df2)[2] <- "Studieschuld_UK"
colnames(df3)[1] <- "Year"
colnames(df3)[2] <- "Studieschuld_US"

#Year was blijkaar numeriek dus een 'karakter' van maken
df2$Year <- as.character(df2$Year)
df3$Year <- as.character(df3$Year)

#Stond een * bij 2023 en 24 bij NL waardoor hij dat als aparte variabele zag
df1$Year[df1$Year == "2023*"] <- 2023
df1$Year[df1$Year == "2024*"] <- 2024

# df1,2,3 samenvoegen
df_merge <- full_join(df1, df2, by = "Year")
df_merge <- full_join(df_merge, df3, by = "Year")

# Sorteren van Oud naar Nieuw
df_sorted <- df_merge[order(df_merge$Year, decreasing = FALSE), ]


# Alles numeriek maken
df_sorted$Studieschuld_NL <- as.numeric(df_sorted$Studieschuld_NL)
df_sorted$Studieschuld_UK <- as.numeric(df_sorted$Studieschuld_UK)
df_sorted$Studieschuld_US <- as.numeric(df_sorted$Studieschuld_US)

# Nieuwe Variabele: Jaarlijkse groei berekenen (percentage)
df_sorted$Growth_NL <- c(NA, diff(df_sorted$Studieschuld_NL) / head(df_sorted$Studieschuld_NL, -1) * 100)
df_sorted$Growth_UK <- c(NA, diff(df_sorted$Studieschuld_UK) / head(df_sorted$Studieschuld_UK, -1) * 100)
df_sorted$Growth_US <- c(NA, diff(df_sorted$Studieschuld_US) / head(df_sorted$Studieschuld_US, -1) * 100)

#Kolommen omzetten in numeriek
df_numeric <- df_sorted %>%
  mutate(across(c(Studieschuld_NL, Studieschuld_UK, Studieschuld_US, Growth_NL, Growth_UK, Growth_US), ~as.numeric(.)),
         Year = as.numeric(Year))

#Gemiddelde growth uitrekenen
NL_Mean_Growth = mean(df_sorted$Growth_NL, na.rm = TRUE)
UK_Mean_Growth = mean(df_sorted$Growth_UK, na.rm = TRUE)
US_Mean_Growth = mean(df_sorted$Growth_US, na.rm = TRUE)


#Alles in Euros

df_numeric <- df_numeric %>%
  mutate(Studieschuld_US = Studieschuld_US * 0.86)

df_numeric <- df_numeric %>%
  mutate(Studieschuld_UK = Studieschuld_UK * 1.17)


# ---- EXTRA GRAFIEK ? ----

#Grafiek
ggplot(df_numeric, aes(x = Year)) +
  geom_line(aes(y = Studieschuld_NL, color = "NL"), size = 1.5) +
  geom_line(aes(y = Studieschuld_UK, color = "UK"), size = 1.5) +
  geom_line(aes(y = Studieschuld_US, color = "US"), size = 1.5) +
  geom_point(aes(y = (Studieschuld_NL), color = "NL"), size = 3) +
  geom_point(aes(y = (Studieschuld_UK), color = "UK"), size = 3) +
  geom_point(aes(y = (Studieschuld_US), color = "US"), size = 3) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "darkred") + 
  annotate(
    "text",
    x = 2023,
    y = 12500,
    label = paste0("Mean growth NL: ", round(NL_Mean_Growth, 2), "%"),
    color = "black",
    fontface = "bold",
    size = 3
  ) +
  annotate(
    "text",
    x = 2023,
    y = 58000,
    label = paste0("Mean growth UK: ", round(UK_Mean_Growth, 2), "%"),
    color = "black",
    fontface = "bold",
    size = 3
  ) +
  annotate(
    "text",
    x = 2023,
    y = 47000,
    label = paste0("Mean growth US: ", round(US_Mean_Growth, 2), "%"),
    color = "black",
    fontface = "bold",
    size = 3
  ) +
  annotate(
    "text",
    x = 2020.7,
    y = 2500,
    label = paste0(" Start Covid-19"),
    color = "black",
    fontface = "bold",
    size = 3
  ) +
  
  scale_x_continuous(
    limits = c(2014, NA),
    breaks = seq(2014, max(as.numeric(df_numeric$Year), na.rm = TRUE), by = 1)  # alle jaren
  ) + 
  labs(title = "Growth in Student Debt per Country",
       x = "Year",
       y = "Student Debt",
       color = "Country") +

  scale_y_continuous(
    limits = c(0, 60000),
    expand = c(0, 0),
    labels = label_dollar(prefix = "€", big.mark = ".", decimal.mark = ",")
  )

  
  theme_bw()

#Opslaan als PNG
ggsave("Temporal_Visualization.png", width = 8, height = 5) 



# ---- SUB POPULATION ----

data_selected <- df_numeric %>%
  select(Year, starts_with("Studieschuld"))

# Data omvormen naar long format
data_long <- data_selected %>%
  pivot_longer(
    cols = -Year,
    names_to = c("variabele", "land"),
    names_pattern = "(Studieschuld)_(.*)"
  ) %>%
  rename(studieschuld = value)

# Tijdsperiode labelen
data_long <- data_long %>%
  mutate(
    period = case_when(
      Year <= 2019 ~ "2014 - 2019",  
      Year > 2019 ~ "2020 - 2024"
    )
  )


# Gemiddelde per land per periode berekenen
data_avg <- data_long %>%
  group_by(land, period) %>%
  summarise(studieschuld_mean = mean(studieschuld, na.rm = TRUE), .groups = "drop")

# Boxplot
ggplot(data_avg, aes(x = period, y = studieschuld_mean)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs(title = "Mean Student Debt per Period (All Countries)",
       x = "Period",
       y = "Mean Student Debt") +
  scale_y_continuous(labels = label_dollar(prefix = "€", big.mark = ".", decimal.mark = ",")) +
  theme_bw()

ggsave("Sub_Population.png", width = 8, height = 5) 



# # Alleen per periode groep maken (alle landen en jaren samen)
# data_period <- data_long %>%
#   group_by(period) %>%
#   summarise(
#     mean_studieschuld = mean(studieschuld, na.rm = TRUE),
#     sd_studieschuld = sd(studieschuld, na.rm = TRUE),
#     min_studieschuld = min(studieschuld, na.rm = TRUE),
#     max_studieschuld = max(studieschuld, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# 
# ggplot(data_long, aes(x = period, y = studieschuld)) +
#   geom_boxplot(fill = "lightblue", color = "darkblue") +
#   labs(title = "Student Debt Distribution per Period (All Countries and Years)",
#        x = "Period",
#        y = "Student Debt") +
#   scale_y_continuous(labels = scales::label_dollar(prefix = "€", big.mark = ".", decimal.mark = ",")) +
#   theme_bw()




# ---- SPATIAL VISUALIZATION ----

#Gemiddelden 
gemiddelde <- data_selected %>% 
  summarise(
    Studieschuld_NL = mean(Studieschuld_NL, na.rm=TRUE),
    Studieschuld_UK = mean(Studieschuld_UK, na.rm=TRUE),
    Studieschuld_US = mean(Studieschuld_US, na.rm=TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "land", values_to = "gemiddelde_schuld") %>%
  mutate(
    land = recode(land,
        Studieschuld_NL = "Netherlands",
        Studieschuld_UK = "United Kingdom",
        Studieschuld_US = "United States of America"
    )
  )


# Wereldkaart
wereldkaart <- ne_countries(scale = "medium", returnclass = "sf")


wereldkaart_met_data <- wereldkaart %>%
  left_join(gemiddelde, by = c("admin" = "land"))


ggplot(wereldkaart_met_data) +
  geom_sf(aes(fill = gemiddelde_schuld), color = "black") +
  scale_fill_gradient(
    low = "lightblue", 
    high = "darkblue", 
    na.value = "lightgrey",      # <- grijs voor landen zonder data
    name = "Mean Student Debt (€)",
    labels = label_dollar(prefix = "€", big.mark = ".", decimal.mark = ",")
  ) +
  coord_sf(
    crs = 3857,
    xlim = c(-1.35e7, 0.4e6),   # links (VS) tot rechts (NL)
    ylim = c(2.2e6, 8e6)       # onder (VS zuid) tot boven (UK noord)
  ) +

  labs(title = "Mean Student Debt (2007–2024)") +  
  theme_bw()



ggsave("Spatial_Visualization.png", width = 8, height = 5) 

# ---- TEMPORAL VISUALIZATION ----

#Procentuele groei per jaar
df_growth_long <- df_numeric %>%
  select(Year, Growth_NL, Growth_UK, Growth_US) %>%
  pivot_longer(
    cols = starts_with("Growth_"),
    names_to = "land",
    names_prefix = "Growth_",
    values_to = "groei_percentage"
  )

df_growth_long %>%
  filter(Year >= 2015) %>%
  ggplot(aes(x = Year, y = groei_percentage, color = land)) +
  geom_line(size = 1.2) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  scale_x_continuous(breaks = 2015:2024) +
  theme_bw() +
  labs(
    title = "Procentual Growth of Student Debt (Since 2014)",
    x = "Year",
    y = "Growth compared to last year (%)",
    color = "Country"
  )

ggsave("Temporal_Visualization2.png", width = 8, height = 5) 


# ---- EVENT ANALYSIS ----

#GDP per capita data inlezen, cleanen en mergen
gdp_per_capita_data <- read_xlsx("gdp_per_capita2.xlsx")


jaar_kolommen <- as.character(unlist(gdp_per_capita_data[3, ]))

jaar_kolommen[is.na(jaar_kolommen)] <- paste0("V", which(is.na(jaar_kolommen)))
df_gdp <- gdp_per_capita_data[4:6, ]
colnames(df_gdp) <- jaar_kolommen
df_gdp <- df_gdp %>%
  select(`Country Name`, `2007`:`2023`)

df_gdp_long <- df_gdp %>%
  pivot_longer(
    cols = -`Country Name`,
    names_to = "Year",
    values_to = "GDP_PC"
  )

df_gdp_long$GDP_PC <- as.numeric(df_gdp_long$GDP_PC)
df_gdp_long$Year <- as.integer(df_gdp_long$Year)

# Vervolgens naar breed formaat
df_gdp_wide <- df_gdp_long %>%
  pivot_wider(
    names_from = `Country Name`,
    values_from = GDP_PC
  ) %>%
  rename(
    GDP_PC_UK = `United Kingdom`,
    GDP_PC_NL = Netherlands,
    GDP_PC_US = `United States`
  ) %>%
  arrange(Year)



#Omzetten van Dollars naar Euro's
df_gdp_wide <- df_gdp_wide %>%
  mutate(GDP_PC_UK = GDP_PC_UK * 0.86)

df_gdp_wide <- df_gdp_wide %>%
  mutate(GDP_PC_NL = GDP_PC_NL * 0.86)

df_gdp_wide <- df_gdp_wide %>%
  mutate(GDP_PC_US = GDP_PC_US * 0.85)



#Samenvoegen met alle andere data en nieuwe variabele maken
df_merged <- full_join(df_numeric, df_gdp_wide, by = "Year")

df_merged <- df_merged %>%
  mutate(
    Schuld_GDP_NL = 100 * Studieschuld_NL / GDP_PC_NL,
    Schuld_GDP_UK = 100 * Studieschuld_UK / GDP_PC_UK,
    Schuld_GDP_US = 100 * Studieschuld_US / GDP_PC_US
  )


#Plot maken
df_long <- df_merged %>%
  select(Year, Schuld_GDP_NL, Schuld_GDP_UK, Schuld_GDP_US) %>%
  pivot_longer(
    cols = starts_with("Schuld_GDP"),
    names_to = "Land",
    values_to = "Schuld_GDP"
  ) %>%
  mutate(
    Land = case_when(
      Land == "Schuld_GDP_NL" ~ "NL",
      Land == "Schuld_GDP_UK" ~ "UK",
      Land == "Schuld_GDP_US" ~ "US"
    )
  )

df_long %>%
  filter(Year >= 2015, Year <= 2023) %>%
  ggplot(aes(x = Year, y = Schuld_GDP, color = Land)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "darkred") +
  labs(
    title = "Student Debt as percentage of GDP per capita",
    x = "Year",
    y = "Student Debt / GDP per capita (%)",
    color = "Country"
  ) +
  annotate(
    "text",
    x = 2020.7,
    y = 6.250,
    label = paste0("Start Covid-19"),
    color = "black",
    fontface = "bold",
    size = 3
  ) +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1),
      limits = c(0, 150),
      expand = c(0, 0),
      breaks = seq(0,150,25)
    ) +
  scale_x_continuous(breaks = 2014:2023) 
  theme_bw()

ggsave("Event_Analysis.png", width = 8, height = 5) 
